Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-09 paper.tex
updates

\title{Wildcard: End user modification of web applications using a live
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own customizations.

In this paper, we present a prototype of Wildcard, an extension
framework embedded in the browser that empowers users to casually tweak
web applications without needing to program in Javascript or directly
interact with the DOM.

Wildcard shows the data from a web page in a table, and maintains a live
bidirectional connection between the table and the original page. By
directly manipulating the table, users can perform a wide variety of
modifications: sorting/filtering content, adding private annotations,
using spreadsheet formulas to fetch data from other web services, adding
custom UI elements to edit form data, and more.

We present examples of using Wildcard to solve real world problems, and
explain the design principles behind the prototype. In the future, we
envision growing Wildcard into a fully deployed system that helps make
the web into a more malleable medium for all its users.
\end{abstract}

@@

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is both
motivated and skilled they might even be able to implement their own
fix. But for most people in most situations, the only recourse is to
complain to the developers and pray that someone listens---or more
likely, to simply give up. While many have become used to this status
quo, we see it as a tremendous waste of the openness of the Web
platform. Back in 1977, in \emph{Personal Dynamic Media}
\autocite{kay1977}, Alan Kay had originally envisioned personal
computing as a medium that let a user ``mold and channel its power to
his own needs,'' but today's software is more like concrete than clay.

In this paper, we introduce Wildcard, an extension framework embedded in
the browser that lets users tweak web applications without programming
in Javascript or interacting with the DOM. Wildcard shows a structured
table view of the main data in a webpage, which maintains a live
bidirectional connection to the original UI: when the user manipulates
the table, the UI is instantly modified, and vice versa. This simple
framework allows a wide range of possible modifications, ranging from
sorting lists to fetching related data from other websites.

The system builds on two insights from prior work. The first insight is
that a spreadsheet can serve as the backend for a GUI application, which
enables an end user to build applications in an extremely accessible
environment without doing traditional programming. The second insight is
that web scraping---extracting structured data from an unstructured web
page---unlocks value by making it possible to compute over that data.
Our contribution in this work is to combine those two ideas in a novel
way. By extracting structured data from an application and showing that
data in a live spreadsheet table, Wildcard enables a wide class of end
user modifications to existing applications. Wildcard makes it seem like
an application is backed by a spreadsheet, even though the actual
implementation is not. This allows the user to easily manipulate and
compute over the application's underlying data just as they would in a
spreadsheet, and then to immediately see the results of their changes in
the original application's UI.

Wildcard embodies several principles we propose for designing systems
that enable end user modification of existing software:

\emph{Expose consistent structure}: Most applications only expose a user
interface, without showing the structure of the data underneath.
Furthermore, every application has a different interface, requiring
users to learn many ways of using apps. Wildcard shows data from diverse
applications in a single consistent structure, so that users are able to
invest in learning one mechanism for manipulating the data in their
applications, and for programming with that data.

\emph{Low floor, high ceiling}: We aim to provide a low barrier to entry
so that non-programmers can casually use the system, but a high ceiling
on the complexity it can achieve. Small but genuinely valuable tweaks
like sorting a page are made possible with little effort in the flow of
normal use to provide a low floor, and a rich formula system creates a
high ceiling on possible modifications.

\emph{Build for multiple tiers of users}: We aim to make Wildcard as
easy as possible for end users, by designing roles for tiers of users
with differing levels of technical sophistication. Programmers or
tech-savvy end users are responsible for the data extraction step of
mapping existing UIs to structured data. Most end users only interact
with the structured data, providing a predictable and user-friendly
experience.

Wildcard is currently an early research prototype, with incomplete
features and limited coverage of sites. We plan to continue privately
testing the system with our own use cases, and then to release the tool
publicly, enabling other programmers to contribute site adapters and
other plugins. Ultimately, we hope that Wildcard makes modifying
websites into a natural part of using the web, as ubiquitous as using
spreadsheets in everyday computing.

@@

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely and declaratively express scraping
logic. For example, here is a code snippet for extracting the name of an
Airbnb listing. It specifies the field name and some metadata
properties, and then a function that, given the entire listing, returns
the DOM element representing the name of the listing.

\begin{Shaded}
\begin{Highlighting}[]
  \OperatorTok{\{}
\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
  \OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is site-specific
actions (like favoriting in Airbnb): actions are implemented as regular
Javascript functions running in the context of the page, which can
mutate the DOM, simulate clicks on buttons in the UI, launch AJAX
requests, navigate to new pages, etc.

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We think these principles extend
beyond Wildcard and might inform the design of other end user
programming tools, particularly systems enabling users to modify GUI
apps.

\hypertarget{expose-consistent-structure}{%
\subsection{Expose consistent
structure}\label{expose-consistent-structure}}

Today, most personal computing consists of using distinct apps, each one
an isolated silo of data and functionality. While there are some shared
idioms and limited points of interoperability, apps fundamentally
operate independently from one another. Computing does not need to
operate this way---for example, UNIX offers a compelling alternate
vision: many small tools, all built on the consistent abstraction of
text files. Users can invest in deeply mastering a text editor, and then
use it for a vast array of tasks, even as an interactive input mechanism
in shell programs (e.g.~for editing git commit messages).
Beaudouin-Lafon and Mackay call this style of interaction
\emph{polymorphic} \autocite{beaudouin-lafon2000}, since a single
interface widget can be used in many contexts. diSessa
\autocite{disessa2000} has also noted the deep connection between
polymorphism and literacy: textual literacy rests on a single rich
medium of writing, which once mastered, can be adapted to many different
genres and uses. If people needed to relearn from scratch when switching
from writing essays to writing emails, the medium would lose most of its
potency.

With Wildcard, we hope to make isolated Web applications work more like
UNIX, by imposing a consistent and simple data structure onto many
applications. Just as UNIX uses text files to represent all data,
Wildcard uses relational tables. This abstraction strikes a balance
between being simple and generic: a data table is much simpler than the
DOM tree that describes all the details of the UI, but is also generic
enough to describe the essence of many different interfaces.

\begin{figure}
\centering
\includegraphics{media/structure.png}
\caption{Wildcard layers the consistent structure of a data table over
different isolated applications.}
\end{figure}

Exposing structure gives users more choice When UI widgets can compete
on their merits rather than having a monopoly on a given data silo or
application, it creates stronger competition at the interface level. For
example, there is competition among email clients which consume an open
protocol, but not among Facebook or Twitter clients. Exposing structure
also leads to improved UI consistency across apps: UNIX users can reuse
their text editor everywhere, but Web users must use a different rich
text editor for each site. Tim Berners-Lee has written in more detail
about these benefits \autocite{berners-lee2018} in the context of the
SOLID project, which envisions giving users control of their own data as
a mechanism for decoupling data from interfaces (e.g., giving users a
choice of which client to use to consume a given social media feed).
Wildcard has overlapping goals with SOLID, but does not require
decentralized user control of data---the data can remain on a
centralized server, as long as the interface can be tweaked by end
users.

\emph{todo: move the sloppy contrast, and SOLID, down into related work}

One key challenge of making a structured view successful is ensuring a
close mapping in the user's mind between the new representation and the
original page. Wildcard provides live visual cues as the user navigates
the data table (similar to the highlighting provided by DOM inspectors
in browser developer tools). In our own usage, we have found that this
live highlighting makes it very clear how the two representations map to
each other.

@@

Seymour Papert advocated for programming systems with a ``low floor''
and a ``high ceiling''---making it easy for novices to get started, but
also providing possible a large range of possibilities for more
sophisticated users \autocite{resnick2016}. Unfortunately, many
programming systems only provide either one or the other
\autocite{myers2000}.

\begin{itemize}
\tightlist
\item
  tweaking vs creation from scratch
\item
  in-place toolchain
\item
  easy things easy
\item
  easy things valuable
\item
  this seems clearly justified by the demos
\end{itemize}

rewrite\ldots{}

\begin{longtable}[]{@{}llc@{}}
\toprule
\endhead
& \emph{In-place} & \emph{Not in-place}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser dev tools & editing open source
software\tabularnewline
\bottomrule
\end{longtable}

We can evaluate a system for tweaking software along two dimensions.
First, the technical capability required of the user: is programming
knowledge needed? Second, whether the change is ``in-place''
\autocite{inkandswitch2019}: can it be made within the same environment
that the software is typically used in? These dimensions are not
orthogonal, but they are distinct. For example, setting up a workflow
trigger in an end user programming system like
\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
but it does require leaving the user's normal software and entering a
separate environment. On the other hand, running a Javascript snippet in
the browser console requires programming skills, but can be done
immediately and casually in the flow of using a website.

In addition to requiring no programming skills, Wildcard supports
frequent, small modifications. The Wildcard table appears in the course
of normal web browsing, to ensure that the tools for modification are
close at hand while using the original software.Wildcard also provides
live feedback, immediately editing the page in response to user
interaction, which gives users confidence in the changes they are
making.

Spreadsheets can be used in genuinely useful ways even by someone who
has learned only a tiny part of their functionality, e.g.~for storing
tables of numbers or computing simple sums \autocite{nardi1991}. This
creates inherent motivation to continue using the tool and eventually
learn its more powerful features. In contrast, many traditional
programming systems require an enormous time investment before a
beginner is able to create any valuable software. Like spreadsheets,
Wildcard aims to support genuinely useful actions with little effort:
for example, sorting a table with a single click, or adding an
annotation by typing into a cell. We would expect many Wildcard users to
start by using these bits of functionality, before possibly moving on to
more complex features like formulas.

High ceiling?

@@

\begin{itemize}
\tightlist
\item
  make things easy for the last tier of users (low floor)
\item
  expose predictable structure, don't make them participate in data
  extraction
\end{itemize}

Some end user programming systems are designed with a single user in
mind---a lone non-programmer, needing to achieve a task through some
combination of individual skills and a powerful tool. But in fact,
real-world usage of end user programming systems is far more
collaborative. Spreadsheets are built together by users with different
levels of technical expertise: some users just write simple formulas,
while their officemates help write more complex formulas or even program
macros in traditional languages \autocite{nardi1990}. Our goal with
Wildcard is to consider the full ecosystem of users and to design a tool
that best combines all of their abilities.

This design principle is most salient in our approach to building site
adapters. Many web augmentation projects aim to make the full process of
tweaking available to end users with no programmer help needed, through
a combination of automated heuristics and interactive feedback. This is
a laudable goal because it makes these tools work with any site, and
these approaches are often quite successful, but they also inevitably
break down sometimes \autocite{bolin2005} \autocite{little2010}. In
contrast, human programmers working together have been able to maintain
stable scraping libraries, even for complex sites like Gmail
\autocite{streak,talwar2019}. Our primary approach to building site
adapters is having programmers manually create them, because we are
willing to trade off lower site coverage in return for more reliable
results on the sites that are covered, building confidence for users.
Like the collaboration between an expert and novice spreadsheet user,
this architecture leverages more skilled programmers while still
providing end users with a flexible programming environment to meet
their own needs. On the other hand, we remain very open to incorporating
other tools in the future that enable end users to also create their own
adapters.

\begin{figure}
\centering
\includegraphics{media/adapters.png}
\caption{Having third party programmers create site adapters balances
site coverage and adapter quality.}
\end{figure}

In any system for tweaking software, another important consideration is
how to involve the first party developers of the original software. Our
approach is to avoid depending on their cooperation, but also to aim for
eventual cooperation.

This approach takes advantage of the unusual openness of the Web
platform. On many other platforms (e.g.~smartphone operating systems),
software is locked down unless first-party developers explicitly provide
hooks for plugins and interoperability, but on the Web, all client-side
code is available for browser extensions to modify. Application authors
can use practices that make it easier to modify their apps (e.g.~clean
semantic markup), or more difficult (e.g.~code obfuscation), but the
default state is openness. This gives extensions freedom to modify
applications in creative ways that the original developers did not plan
for. At the most extreme, this can even include modifications that are
hostile to the desires of the original developers. Although ad blocking
extensions make this style of extension very prominent, most ideas for
extending software merely fall outside the range of possibilities
originally imagined by the developer, rather than being actively
hostile.

Wildcard takes advantage of this openness, and does not depend on
cooperation from first-party website developers: any programmer can add
support for any website to Wildcard by building a third party adapter.
But the platform is still open to first parties developing their own
adapters, which would often be in their interests. Providing Wildcard
support would allow users to build extensions to fulfill their own
feature requests. It also would not necessarily require much effort:
adding Wildcard support would be more straightforward for a first-party
than a third-party because they have direct access to the structured
data in the page. There is also precedent for first parties implementing
an official client extension API in response to user demand: for several
years, Google maintained an official extension API in Gmail for
Greasemonkey scripts to use. (Incidentally, since then, third parties
have continued to maintain stable Gmail extension APIs used by many
browser extensions \autocite{streak,talwar2019}, illustrating the
potential of collaboratively maintaining third party adapters.)

@@

\hypertarget{web-augmentation}{%
\subsection{Web augmentation}\label{web-augmentation}}

Wildcard's goals are closely shared with other systems that provide
interfaces in the browser for end users to augment and modify websites
while using them.

@@

>>>>

\title{Wildcard: Spreadsheet-Driven Customization of Web Applications}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own extensions.

We present the idea of \emph{spreadsheet-driven customization}: enabling
end users to customize existing applications using a live spreadsheet
view of the data inside the application. By manipulating the
spreadsheet, users can implement a wide variety of customizations,
ranging from sorting lists of search results to displaying related data
from other web services, without doing any traditional programming.

We built a prototype system called Wildcard that implements
spreadsheet-driven customization as a web browser extension. Through
concrete examples, we demonstrate that Wildcard has both a low barrier
to entry for beginners and enough flexibility to solve many useful
problems. We also show that Wildcard can work with real existing
websites, by extracting structured data using web scraping techniques.

For an online version of this paper with videos demonstrating the interface, see \url{https://www.geoffreylitt.com/wildcard/}
\end{abstract}

@@

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is both
motivated and skilled they might even be able to implement their own
fix. But for most people, the only recourse is to complain to the
developers and pray that someone listens---or more likely, to simply
give up. While many have become used to this status quo, we see it as a
tremendous waste of the openness of the Web platform. Back in 1977, in
\emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay originally
envisioned personal computing as a medium that let a user ``mold and
channel its power to his own needs,'' but today software feels more like
concrete than clay.

In this paper, we present \emph{spreadsheet-driven customization}, a
technique for making software customization more accessible to end
users. The core idea is to show the data inside the application in a
spreadsheet which maintains a live connection to the original UI. When
the user manipulates the spreadsheet, the UI is instantly modified, and
vice versa. We built a research prototype called Wildcard that
implements spreadsheet-driven customization as a web browser extension,
showing a spreadsheet panel in the context of existing sites.

Spreadsheets have proven to be a widely successful computing platform
for non-programmers. They provide a low barrier to entry for beginners,
while allowing experts to perform complex computations. The 2D grid is
simple to grasp, but general enough to represent many types of data.
Prior work \autocite{mccutchen2016,benson2014,chang2014} has used these
benefits to enable end users to easily create web applications that use
spreadsheets as a backing data layer.

Spreadsheet-driven customization applies these exact same benefits in a
different context: customizing existing software, rather than building
new software from scratch. We inherit the low floor and high ceiling of
spreadsheets: small tweaks like sorting a list of data in an app can be
done with a single click, while more complex customizations, like
joining in related data from a web API, are supported by a rich formula
system. The tabular data format is versatile, supporting data from a
wide range of applications.

Spreadsheet-driven customization does not require that the application
is actually backed by a spreadsheet---it merely presents a model of the
internal state of the application, which can be exposed by the
application itself or derived from the user interface by third parties.
This means that spreadsheet-driven customization works with existing
applications that people already use. It also allows for exposing a
common data abstraction across applications, enabling users to have a
consistent mechanism for manipulating data from all of their
applications.

Prior work \autocite{huynh2006} has shown that the process of extracting
structured data from existing user interfaces can be confusing and
unfamiliar for end users. In order to make the customization experience
as straightforward as possible, we envision a tiered architecture that
hides this complexity from end users. Wildcard provides a mechanism for
programmers to write \emph{site adapters}, which use web scraping
techniques to extract structured data from the DOM and AJAX requests of
existing applications. End users only interact with the structured
spreadsheet, providing a predictable experience with clear affordances
for which customizations are possible.

Wildcard is currently an early prototype, with incomplete features and
limited coverage of sites. We plan to continue building site adapters
and testing the system with our own use cases, to better understand how
well the spreadsheet abstraction maps to real websites and customization
needs. Eventually we also plan to release the tool publicly, to
determine how end users choose to use spreadsheet-driven customization,
what usability challenges emerge, and how feasible it is for programmers
to build and maintain site adapters.

@@

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely express web scraping logic. For
example, here is a code snippet for extracting the name of an Airbnb
listing:

\begin{Shaded}
\begin{Highlighting}[]
  \OperatorTok{\{}
\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the data field}
\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}
\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
    \CommentTok{// Given the DOM element for the entire listing,}
    \CommentTok{// return the DOM element representing this field}
\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
  \OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

Scraping data out of the UI has the advantage of working for any data
visible on the page, but it can require a lot of work to convert
unstructured UI to structured data, and sometimes important data is not
shown in the UI. To work around these problems, we have also prototyped
mechanisms for site adapters to observe AJAX requests made by the
browser and extract data directly from JSON responses. (This mechanism
was used in the Airbnb Walkscore example: approximate latitude and
longitude aren't shown in the Airbnb UI, but they are available in AJAX
responses.)

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is row actions (like
favoriting an Airbnb listing), which are implemented as regular
Javascript functions running in the context of the page and can mutate
the DOM, simulate clicks on buttons, etc.

@@

The idea of spreadsheet-driven customization is guided by several design
principles, inspired by prior work and our own experimentation. We think
these principles can also broadly inform the design of end user
programming tools, especially those that enable users to customize
existing software.

\hypertarget{expose-a-universal-data-structure}{%
\subsection{Expose a universal data
structure}\label{expose-a-universal-data-structure}}

Today, most personal computing consists of using applications, which
bundle together behavior and data to provide some set of functionality.
While there are limited points of interoperability, applications
generally are designed to operate independently of one another. Once a
user has gotten past learning the basic idioms of modern computing like
windows and scrollbars, most of the effort invested in learning to use
one application does not carry over to using other applications.

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
utilities, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
can get a lot of utility from deeply mastering a text editor and some
text manipulation utilities, because these tools can be applied to
nearly any task. As just one example, a user's preferred text editor can
even serve as an interactive input mechanism in shell programs, e.g.~for
editing git commit messages.

Spreadsheet-driven customization aims to port this UNIX philosophy to
the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table: a simple abstraction generic enough to describe the data used in
many different applications. Because Wildcard maps the data from all
applications to the table format, users can invest in mastering the
Wildcard table editor, the formula language, and cell editor UIs, and
reuse those same tools to customize many different applications.

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000}:
UI elements that can be used in different contexts, like a color picker
that can be used in many different drawing applications. diSessa has
also noted the connection between literacy and the genericness of a
medium. Textual literacy rests on the fact that writing can be adapted
to many different genres and uses \autocite{disessa2000}; if people
needed to relearn reading and writing from scratch when switching from
essays to emails, the medium would lose most of its potency. We think
providing generic tools is especially important for software
customization, because the most common barrier to customizing software
is not having enough time \autocite{mackay1991}---it's more likely that
people will customize software regularly if they can reuse the same tool
across many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and can't necessarily naturally express
the data in every application. We plan to explore the limits of the
table abstraction further, by trying to build adapters for more sites
with varied data formats. We expect that many types of data can fit
fairly naturally into tables: lists of search results, news articles,
and messages can all be seen as relations. On sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social friend
graphs), it may prove more challenging to map internal data to this
abstraction.

Another challenge is ensuring a clear mapping in the user's mind between
the spreadsheet and the original page. Wildcard provides live visual
cues as the user navigates the data table (similar to the highlighting
provided by DOM inspectors in browser developer tools). In our own
usage, we have found that this live highlighting makes it very clear how
the two representations map to each other.

@@

Seymour Papert advocated for programming systems with a ``low floor,''
making it easy for novices to get started, as well as a ``high
ceiling,'' providing a large range of possibilities for more
sophisticated users \autocite{resnick2016}. Our goal is for
spreadsheet-driven customization to meet both of these criteria.
Although we need more testing with real users to know for sure, we think
Wildcard clearly provides a low floor, and likely has a flexible enough
foundation to eventually provide a high ceiling.

One of the most interesting properties of spreadsheets is the amount of
value they can provide to users who are aware of only a tiny sliver of
their functionality. Not only can a novice perform some basic activities
in a spreadsheet with almost no training (e.g., storing tables of
numbers or computing simple sums), these activities are often actually
valuable for the user! The fact that useful tasks can be performed early
on supports the user's natural motivation to continue using the tool,
and to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront investment of time and practice before
someone is able to write a program that actually helps them achieve a
real task.

As part of ensuring a low floor for spreadsheet-driven customization, we
have focused on including genuinely valuable features for novices. For
example, a user can sort a table with a single click, or simply type in
some annotations. We would expect many Wildcard users to start out using
these simpler features before potentially moving on to more
sophisticated features like formulas.

Another aspect of providing a low floor is providing an ``in-place
toolchain'' \autocite{inkandswitch2019}---minimizing the effort of
moving from using to customizing, by making customization tools
available in the same environment where the user is already using the
software. This quality is distinct from the level of technical skill
needed to use the tool: for example, setting up a workflow trigger in an
end user programming system like \href{https://ifttt.com/}{IFTTT} does
not require much technical skill, but does require leaving the user's
normal software and entering a separate environment; conversely, running
a Javascript snippet in the browser console requires programming skills,
but can be done immediately and casually in the flow of using a website.

\begin{longtable}[]{@{}llc@{}}
\toprule
\endhead
& \emph{In-place} & \emph{Not in-place}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser JS console & forking open source
software\tabularnewline
\bottomrule
\end{longtable}

Wildcard provides an in-place toolchain---on any site that supports
Wildcard, the user is one click away from starting to tweak the site
with minimal friction. There is no need for them to switch into another
environment or figure out how to extract data from the page. Once the
user starts editing, Wildcard also provides live feedback to help users
understand the changes they are making. Even if a user isn't yet totally
familiar with Wildcard, they can learn to use the system by trying
things out and seeing what happens.

Since we have still only built several site adapters and demos, it's
still too early to tell how high the ceiling is for the customizations
that can be achieved with Wildcard, but we think that with enough
operators the formula language could eventually support a wide variety
of customizations. We plan to explore this aspect further by trying to
solve more real problems with the system and observing where limitations
emerge in practice.

@@

Real-world spreadsheet usage in offices is highly collaborative: most
users just perform simple changes, while a few coworkers help with
writing more complex formulas or even programming macros
\autocite{nardi1990}. Inspired by this, we aim to make
spreadsheet-driven customization a collaborative activity that combines
the different abilities of many users in a collaborative ecosystem.

The main way we do this is by separating website customization into two
separate stages: structured data extraction, and using the resulting
spreadsheet. The first stage is currently only available to programmers
who can code site adapters in Javascript, whereas the second stage is
available for any non-programmer end user. This architecture frees end
users from needing to think about data extraction, and enables a
community of end users to reuse the efforts of programmers building site
adapters.

The group of users building adapters does not necessarily need to be
limited only to programmers. In the future we might explore enabling end
users to also create site adapters, drawing on related work on enabling
end users to extract structured data from websites
\autocite{chasins2018,huynh2006}. But even in that case, there would
still be a separation between motivated, tech-savvy end users building
adapters and more casual end users just using the spreadsheet view.

Another stakeholder to consider is the first party developers of the
original software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers: a third party
programmer can write an adapter for any website, which can access any
information available in the browser. On the other hand, if first
parties were to expose structured data from their applications, it would
avoid the need for adapters and generally make customization a lot
easier. We think there are compelling reasons for first parties to
consider doing this. Providing Wildcard support would allow users to
build extensions to fulfill their own feature requests. It also would
not necessarily require much effort: adding Wildcard support would be
more straightforward for a first-party than a third-party because they
have direct access to the structured data in the page. There is also
precedent for first parties implementing an official client extension
API in response to user demand: for several years, Google maintained an
official extension API in Gmail for Greasemonkey scripts to
use.\footnote{Incidentally, since then, third parties have continued to
  maintain stable Gmail extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

\hypertarget{web-customization}{%
\subsection{Web customization}\label{web-customization}}

Wildcard's goals are closely shared with other systems that provide
interfaces in the browser for end users to augment and customize
websites while using them.

@@

\emph{Copy of the sloppy contrast that used to be earlier:}
